Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to agroecossystems

### 2021-08-03 Manuscript/Manuscript.Rmd
llkl

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States [@ward2013]. The species has showed a remarkable capacity to evolve resistance to herbicides. Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Uncontrolled Palmer amaranth in competition for water, light and nutrients can drastically reduce crop yields [@berger2015]. Palmer amaranth is documented with potential to reduce 91%, 68%, and 54% of corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Thus, Palmer amaranth poses a risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. Farm mechanization, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@ward2013]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017], and plant propagules [@yu2021]. Therefore, Palmer amaranth dispersal capacity make the species one of the most successful cases of weed adaption to cropping systems.

@@

Palmer amaranth is already documented in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is established at crop [@garetson2019] and non-crop land [@bagavathiannan2016] in the warm southern United States but its range is expanding to cool temperatures northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], and Connecticut [@aulakh2021]. Successful cases of Palmer amaranth invasion and near to eradication is reported in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Therefore, strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management.

The continuous Palmer amaranth dispersal and potential establishment into northern United States warrant investigations on species morphology growing in such environments.  Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020].  Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under in corn, soybean and fallow at two timings across five locations in the mid/upper United States Midwest.

@@

@@

>>>>

Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently considered one of the most economically damaged weed species to cropping systems in the United States . The species has showed a remarkable capacity to evolve resistance to herbicides. To date, Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Unmanaged Palmer amaranth in competition for water, light and nutrients can drastically impact on crop yields [@berger2015]. For example, Palmer amaranth is documented with potential to reduce 91%, 68%, and 54% of corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Thus, Palmer amaranth poses an economical risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with species biology. Farm mechanization, conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017], and plant propagules [@yu2021]. The dispersal capacity of Palmer amaranth make the species one of the most successful cases of weed adaption to cropping systems.

@@

Palmer amaranth is already documented in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the US, Palmer amaranth is well established at crop [@garetson2019] and non-crop land [@bagavathiannan2016] in the warm southern United States but its range is expanding to cool temperatures northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a],  Michigan [@kohrt2017], and Connecticut [@aulakh2021]. Successful cases of Palmer amaranth invasion and near to eradication is reported in Minnesota [@yu2021]. No Palmer amaranth actively growing was found in Canada; however, Palmer amaranth seeds was detected in sweet potato slips [@page2021]. Nonetheless, it seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Therefore, strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Strategies to manage Palmer amaranth should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

The continuous Palmer amaranth dispersal and potential establishment into northern United States warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

For example, Palmer amaranth growing at first cohort accumulated 90%, 85% and 31% more biomass when compared to second cohort at corn, soybean and fallow, respectively

In addition, Palmer amaranth growing in corn and soybean at second cohort had a similar end-season flowering pattern.

@@

Our study showed the Palmer amaranth flowering pattern, grow and development varied within agroecossystems and timing. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. Thus, Palmer amaranth growing strategy was to mimic the crop species (Figure 4). In contrast, at second cohort, Palmer amaranth grow and development was limited due to the crop competitive ability at advanced development stages.

Taller Palmer amaranth plants is a possible strategy to compete for light in between crop rows in absence of canopy. For example, at first cohort, corn and Palmer amaranth had abundant light to grow and develop, which also aid its extended the flowering pattern. However, at second cohort, Palmer amaranth was transplanted when corn canopy was nearly closed, which reduced Palmer amaranth competition. As a result, Palmer amaranth height and biomass was nearly 50% and 90% lower compared to its first cohort, respectively. Similar trend was observed to Palmer amaranth growing in soybeans. Palmer amaranth growing without crop competition produced highest amount of biomass. The Palmer amaranth strategy in fallow was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 31% less biomass in second cohort compared to first cohort timing. Therefore, crop competition was not the only factor limiting Palmer amaranth growing in the second cohort.

It was not evaluate seed production due to Palmer amaranth harvest at early flowering. Nonetheless, it is reported a strong positive correlation between Palmer amaranth biomass and seed production [@schwartz2016; @spaunhorst2018]. I our study, Palmer amaranth growing at first cohort accumulated nearly 50% more overall biomass when compared to second cohort. Therefore, Palmer amaranth plants growing in the second cohort is likely to produce less seeds regardless the agroecossystem. Seed production and deposition in the seedbank is a key factor for species perpetuation [@menges1987]. Palmer amaranth germination is influenced by soil depth and seed viable seeds were fount after 36 months buried in the seedbank [@sosnoskie2013]. Therefore, tactics that promote crop advantage against Palmer amaranth, including early crop planting, preemergence applied herbicide, and crop residue (e.g. cover crops) can minimize the negative impact of Palmer amaranth in agroecossystems.

Palmer amaranth has shown to be responsive to to thermal units (e.g, growing degree days). A study in North Carolina demonstrated that Palmer amaranth emergence in response to soil temperature, with seedlings emerging before July prior to crop canopy closure [@piskackova2021]. In addition, in a bareground study, Palmer emergence emergence pattern in Nebraska was influenced by soil management. Early tillage resulted in 90% Palmer amaranth emergence in June

The higher amount of biomass in the first cohort can also be related to the longer flowering period in the first cohort. In this study, biomass is related to flowering as plant was harvest at early flowering.

can cause significant soybean yield loss [@korres2019]

### 2021-08-03 Manuscript/Manuscript.tex
llkl

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States (Ward et al., 2013). The species has showed
a remarkable capacity to evolve resistance to herbicides. Palmer
amaranth has evolved resistance to eight herbicide sites of action
(Heap, 2021), increasing the weed management complexity (Lindsay et al.,
2017). Uncontrolled Palmer amaranth in competition for water, light and
nutrients can drastically impact on crop yields (Berger et al., 2015).
Palmer amaranth is documented with potential to reduce 91\%, 68\%, and
54\% of corn (Massinga et al., 2001), soybean (Klingaman and Oliver,
1994), and cotton (Morgan et al., 2001) yields, respectively. Thus,
unmanaged Palmer amaranth poses an economical risk to sustainable
agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with species
biology. Farm mechanization, conservation agriculture (e.g., no-till),
and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems. On
the other hand, Palmer amaranth is a prolific seed producer with a C4
photosynthetic apparatus (Ward et al., 2013). With a dioecy nature,
Palmer amaranth male and female plants are obligate outcrosser species,
increasing the chances of exchanging adaptive traits among plants
(Oliveira et al., 2018). Also, Palmer amaranth small seed size (e.g, 1
mm) tend to thrive in no-tillage systems (Price et al., 2011), and
spread across locations through farm equipment (Sauer, 1972), manure
(Hartzler and Anderson, 2016), animals (Farmer et al., 2017), and plant
propagules (Yu et al., 2021). Palmer amaranth dispersal capacity make
the species one of the most successful cases of weed adaption to
cropping systems.

@@

Palmer amaranth is already documented in agronomic crops of South
America (Larran et al., 2017; Küpper et al., 2017) and Southern Europe
(Milani et al., 2021). In the US, Palmer amaranth is well established at
crop (Garetson et al., 2019) and non-crop land (Bagavathiannan and
Norsworthy, 2016) in the warm southern United States but its range is
expanding to cool temperatures northward. For example, herbicide
resistant Palmer amaranth is widespread in Nebraska (Oliveira et al.,
2021), Michigan (Kohrt et al., 2017), and Connecticut (Aulakh et al.,
2021). Successful cases of Palmer amaranth invasion and near to
eradication is reported in Minnesota (Yu et al., 2021). No Palmer
amaranth actively growing was found in Canada; however, Palmer amaranth
seeds was detected in sweet potato slips (Page et al., 2021).
Nonetheless, it seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Therefore, strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Long-term tactics to minimize the negative
Palmer amaranth impact in the agroecossystem should include species
strategies to adapt and grow into different environments.

The continuous Palmer amaranth dispersal and potential establishment
into northern United States warrant investigations on species morphology
growing in such environments. Understanding Palmer amaranth biology and
growing strategies under different agroecossystems can enhance our
knowledge on species adaptation. It can also aid on designing proactive
and ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under in corn, soybean and fallow at two timings
across five locations in the mid/upper United States Midwest.

@@

@@

Palmer amaranth cumulative flowering from first transplant timing ranged
from late June to early September.

@@

@@

@@

@@

@@

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States . The species has showed a remarkable
capacity to evolve resistance to herbicides. To date, Palmer amaranth
has evolved resistance to eight herbicide sites of action (Heap, 2021),
increasing the weed management complexity (Lindsay et al., 2017).
Unmanaged Palmer amaranth in competition for water, light and nutrients
can drastically impact on crop yields (Berger et al., 2015). For
example, Palmer amaranth is documented with potential to reduce 91\%,
68\%, and 54\% of corn (Massinga et al., 2001), soybean (Klingaman and
Oliver, 1994), and cotton (Morgan et al., 2001) yields, respectively.
Thus, Palmer amaranth poses an economical risk to sustainable
agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with species
biology. Farm mechanization, conservation agriculture (e.g., no-till),
and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Oliveira et al., 2018). Also, Palmer amaranth small seed
size (e.g, 1 mm) tend to thrive in no-tillage systems (Price et al.,
2011), and spread across locations through farm equipment (Sauer, 1972),
manure (Hartzler and Anderson, 2016), animals (Farmer et al., 2017), and
plant propagules (Yu et al., 2021). The dispersal capacity of Palmer
amaranth make the species one of the most successful cases of weed
adaption to cropping systems.

@@

Palmer amaranth is already documented in agronomic crops of South
America (Larran et al., 2017; Küpper et al., 2017) and Southern Europe
(Milani et al., 2021). In the US, Palmer amaranth is well established at
crop (Garetson et al., 2019) and non-crop land (Bagavathiannan and
Norsworthy, 2016) in the warm southern United States but its range is
expanding to cool temperatures northward. For example, herbicide
resistant Palmer amaranth is widespread in Nebraska (Oliveira et al.,
2021), Michigan (Kohrt et al., 2017), and Connecticut (Aulakh et al.,
2021). Successful cases of Palmer amaranth invasion and near to
eradication is reported in Minnesota (Yu et al., 2021). No Palmer
amaranth actively growing was found in Canada; however, Palmer amaranth
seeds was detected in sweet potato slips (Page et al., 2021).
Nonetheless, it seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Therefore, strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Strategies to manage Palmer amaranth should
be built on minimizing the species ability to adapt, grow and develop
into agroecossystems.

The continuous Palmer amaranth dispersal and potential establishment
into northern United States warrant investigations on species morphology
in such environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under corn, soybean and fallow at two timings
across five locations in the United States Midwest.

@@

For example, Palmer amaranth growing at first cohort accumulated 90\%,
85\% and 31\% more biomass when compared to second cohort at corn,
soybean and fallow, respectively

In addition, Palmer amaranth growing in corn and soybean at second
cohort had a similar end-season flowering pattern.

@@

Our study showed the Palmer amaranth flowering pattern, grow and
development varied within agroecossystems and timing. In general, Palmer
amaranth produced more biomass and taller plants when growing at first
cohort rather than second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
both species, crop and weed. Thus, Palmer amaranth growing strategy was
to mimic the crop species (Figure 4). In contrast, at second cohort,
Palmer amaranth grow and development was limited due to the crop
competitive ability at advanced development stages.

Taller Palmer amaranth plants is a strategy to compete for light in
between crop rows in absence of canopy. For example, at first cohort,
corn and Palmer amaranth had abundant light to grow and develop, which
likely aid its extended the flowering pattern. However, at second
cohort, Palmer amaranth was transplanted when corn canopy was nearly
closed, which reduced Palmer amaranth competition. As a result, Palmer
amaranth height and biomass was nearly 50\% and 90\% lower compared to
its first cohort, respectively. Similar trend was observed to Palmer
amaranth growing in soybeans. Palmer amaranth growing without crop
competition produced highest amount of biomass. The Palmer amaranth
strategy in fallow was to invest biomass in growing plant width and
height. Nonetheless, Palmer amaranth produced 31\% less biomass in
second cohort compared to first cohort timing. Therefore, crop
competition was not the only factor limiting Palmer amaranth growing in
the second cohort.

It was not evaluate seed production due to Palmer amaranth harvest at
early flowering. Nonetheless, it is reported a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated nearly 50\% more overall
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the agroecossystem. Seed production and deposition in the
seedbank is a key factor for species perpetuation (Menges, 1987). Palmer
amaranth germination is influenced by soil depth and seed viable seeds
were fount after 36 months buried in the seedbank (Sosnoskie et al.,
2013). Therefore, tactics that promote crop advantage against Palmer
amaranth, including early crop planting, preemergence applied herbicide,
and crop residue (e.g.~cover crops) can minimize the negative impact of
Palmer amaranth in agroecossystems.

Palmer amaranth has shown to be responsive to to thermal units (e.g,
growing degree days). A study in North Carolina demonstrated that Palmer
amaranth emergence in response to soil temperature, with seedlings
emerging before July prior to crop canopy closure (Piskackova et al.,
2021). In addition, in a bareground study, Palmer emergence emergence
pattern in Nebraska was influenced by soil management. Early tillage
resulted in 90\% Palmer amaranth emergence in June

can cause significant soybean yield loss (Korres et al., 2019)

@@

\leavevmode\hypertarget{ref-korres2019}{}%
Korres, N. E., Norsworthy, J. K., and Mauromoustakos, A. (2019). Effects
of {Palmer Amaranth} ({Amaranthus} palmeri) {Establishment Time} and
{Distance} from the {Crop Row} on {Biological} and {Phenological
Characteristics} of the {Weed}: {Implications} on {Soybean Yield}.
\emph{Weed Science} 67, 126--135.
doi:\href{https://doi.org/10.1017/wsc.2018.84}{10.1017/wsc.2018.84}.

@@

\leavevmode\hypertarget{ref-menges1987}{}%
Menges, R. M. (1987). Weed {Seed Population Dynamics} during {Six Years}
of {Weed Management Systems} in {Crop Rotations} on {Irrigated Soil}.
\emph{Weed Science} 35, 328--332. Available at:
\url{http://www.jstor.org/stable/4044593}.

@@

\leavevmode\hypertarget{ref-piskackova2021}{}%
Piskackova, T. A. R., Reberg-Horton, S. C., Richardson, R. J., Jennings,
K. M., Franca, L., Young, B. G., and Leon, R. G. (2021). Windows of
action for controlling palmer amaranth ({Amaranthus} palmeri) using
emergence and phenology models. \emph{Weed Research} 61, 188--198.
doi:\href{https://doi.org/10.1111/wre.12470}{10.1111/wre.12470}.

@@

\leavevmode\hypertarget{ref-schwartz2016}{}%
Schwartz, L. M., Norsworthy, J. K., Young, B. G., Bradley, K. W.,
Kruger, G. R., Davis, V. M., Steckel, L. E., and Walsh, M. J. (2016).
Tall {Waterhemp} ({Amaranthus} tuberculatus) and {Palmer} amaranth
({Amaranthus} palmeri) {Seed Production} and {Retention} at {Soybean
Maturity}. \emph{Weed Technology} 30, 284--290.
doi:\href{https://doi.org/10.1614/WT-D-15-00130.1}{10.1614/WT-D-15-00130.1}.

@@

\leavevmode\hypertarget{ref-sosnoskie2013}{}%
Sosnoskie, L. M., Webster, T. M., and Culpepper, A. S. (2013).
Glyphosate {Resistance Does Not Affect Palmer Amaranth} ({Amaranthus}
palmeri) {Seedbank Longevity}. \emph{Weed Science} 61, 283--288.
doi:\href{https://doi.org/10.1614/WS-D-12-00111.1}{10.1614/WS-D-12-00111.1}.

\leavevmode\hypertarget{ref-spaunhorst2018}{}%
Spaunhorst, D. J., Devkota, P., Johnson, W. G., Smeda, R. J., Meyer, C.
J., and Norsworthy, J. K. (2018). Phenology of {Five Palmer} amaranth
({Amaranthus} palmeri) {Populations Grown} in {Northern Indiana} and
{Arkansas}. \emph{Weed Science} 66, 457--469.
doi:\href{https://doi.org/10.1017/wsc.2018.12}{10.1017/wsc.2018.12}.

\leavevmode\hypertarget{ref-wang1992}{}%
Wang, J. L., Klessig, D. F., and Berry, J. O. (1992). Regulation of {C4
Gene Expression} in {Developing Amaranth Leaves}. \emph{The Plant Cell}
4, 173--184.
doi:\href{https://doi.org/10.1105/tpc.4.2.173}{10.1105/tpc.4.2.173}.
